Intra and interspecific diversity in a tropical plant clade alter herbivory and ecosystem resilience

Version 1

Abstract

Abstract

Declines in biodiversity generated by anthropogenic stressors at both species and population levels can alter emergent processes instrumental to ecosystem function and resilience. As such, understanding the role of biodiversity in ecosystem function and its response to climate perturbation is increasingly important, especially in tropical systems where responses to changes in biodiversity are less predictable and more challenging to assess experimentally. Using large scale transplant experiments conducted at five neotropical sites, we documented the impacts of changes in intraspecific and interspecific plant richness in the genus Piper on insect herbivory, insect richness, and ecosystem resilience to perturbations in water availability. We found that reductions of both intraspecific and interspecific Piper diversity had dramatic and site specific effects on herbivory, herbivorous insect richness, and plant mortality. Ecosystem responses to reduced intraspecific richness were often similar in magnitude to responses to reduced interspecific richness. Increased water availability reduced herbivory by 4.2% overall, and the response of herbivorous insect richness and herbivory to water availability was altered by both intra and interspecific richness in a site contingent manner. Our results underscore the role of intraspecific and interspecific richness as foundations of ecosystem function, and the importance of community specific contingencies in controlling function in complex tropical systems.

Introduction

As climate change and anthropogenic activity alter ecosystems at unprecedented rates, it has become critical to understand the consequences of biodiversity loss on ecosystem processes and the maintenance of ecosystem processes through species interactions. A complex mix of anthropogenic forces are eroding multiple dimensions of global biological diversity, including plant intraspecific, interspecific, and functional diversity (1, 2). Plant diversity affects herbivore abundance and diversity, thereby influencing biomass allocation and energy fluxes between trophic levels (3). Because losses of plant diversity can destabilize the flow of resources to higher trophic levels (3, 4), understanding the connection between biodiversity and trophic interactions is necessary to predict the consequences of the loss of primary producer biodiversity on ecosystem traits such as resilience to environmental perturbation (5, 6). While climate change is associated with increased absolute precipitation in some regions and decreased precipitation in others, IPCC models predict mid-century increases in the frequency of extreme precipitation events in Central and South America (7–9), a phenomenon already observable in many ecosystems (10). Measuring how precipitation change will affect the relationship between biodiversity and ecosystem function is therefore increasingly important, particularly in diverse tropical systems.

Very few multi-site, manipulative diversity experiments have been reported from tropical areas compared to temperate environments (11), limiting our knowledge of the role of biodiversity in ecosystem function in the most species-rich regions of the planet (12). Here we investigate how multiple dimensions of plant diversity affect herbivory and plant mortality at five neotropical sites and explore how diversity modulates ecosystem responses to changes in water availability at three of those sites, in the context of several established hypotheses (Table S1). By altering plant uptake of nutrients and the induction of plant defenses, abnormal levels of precipitation can alter herbivore pressure, acting to alter the movement of resources into higher trophic levels. As such, climate extremes of dry and wet periods are expected to strongly perturb plant-insect interactions and thereby alter ecosystem function (13–15). The insurance hypothesis suggests that greater biodiversity can act to stabilize ecosystems and improve their resilience to environmental change (4, 16). Despite traditional views that interspecific richness has a greater impact on ecosystem processes than intraspecific diversity (17), recent research suggests that plant intra- and interspecific richness can have similar effects on ecosystem productivity and consumer abundance (18, 19), and field studies have demonstrated both positive and neutral effects of plant diversity on ecosystem resilience (20, 21).

To test the insurance hypothesis, the effects of water addition on ecosystem function, and the relative strength of intra- and interspecific richness in driving ecosystem function, we conducted common garden experiments at five sites across Central and South America. Using 33 species in the genus Piper (Piperaceae) as a model system (22), we manipulated intra- and interspecific plant richness in Costa Rica, Ecuador, Peru, and two sites in Brazil. We additionally manipulated water availability in Costa Rica, Ecuador, and Peru (Figs. 1 & S1 A). We predicted that reduced Piper diversity would lead to reduced diversity in higher trophic levels, that water addition would lead to altered herbivore pressure, and that lower Piper diversity would be associated with more extreme changes in herbivory and plant mortality in response to water addition. Additionally, we predicted that changes in intraspecific plant richness would affect ecosystem function with magnitudes similar to interspecific richness.

Treatments of intraspecific richness, interspecific richness, and water addition used in each of the five study sites.

White tiles represent treatment combinations which were not tested in a given site.

Methods

Study sites and focal plant genus

We conducted a large-scale transplant experiment replicated across five sites spanning 42° latitude in the Neotropics (Fig. 1 & Table S2 and encompassing a range in annual precipitation from 1271 mm to 4495 mm (Table S2). At each site, we studied herbivory on planted individuals in the genus Piper (Piperaceae) in response to experimental treatments. Study sites included lowland equatorial humid forest at La Selva Biological Station, Costa Rica; high elevation equatorial humid forest in Yanayacu Biological Station, Ecuador; high elevation equatorial humid forest at El Fundo Génova, Peru; lowland seasonally dry gallery forest in cerrado within the phytogeographic domain of the Atlantic Forest in Mogi-Guaçu Biological Reserve, Brazil; and lowland seasonally semideciduous forest in the transition between the Atlantic Forest and the Cerrado phytogeographic domains in Uaimii State Forest, Brazil. Climate classifications follow the Köppen-Geiger climate model (Table S2; 22).

Multiple species of Piper are found at all sites, ranging from 11 species in Mogi-Guaçu to 50 species in Costa Rica (24). Piper is an ideal genus for large-scale comparative studies, as it is found across the Neotropics and subtropics, from ∼10° N to about ∼32° S. In addition to being widespread, Piper is abundant and diverse across its range, encompassing ∼1000 species in the Neotropics (25). Piper has been the subject of detailed studies of herbivory, and its herbivore fauna has been surveyed across its range (26–37). Members of the genus host both specialist caterpillars and beetles as well as generalist caterpillars and Orthopterans (22,26–29,31). Leaf damage patterns produced by the different classes of herbivores are well documented (31)

Experimental design

At each study site, a factorial experiment was implemented to test the effects of plant intraspecific and interspecific richness on herbivory, variation in herbivory, and insect richness. Experimental plants were propagated from cuttings of naturally occurring Piper plants (typically with three nodes and zero to one leaves). In Costa Rica, Ecuador, and Peru a water addition treatment was also tested. In all sites except Uaimii, experimental plots measured 4 m in diameter and contained 12 Piper individuals, planted either in pots with locally derived soil (for experiments manipulating water availability) or directly into the ground (for experiments without water additions in Mogi-Guaçu and Uaimii; Table S2). Plots in Uaimii were 2 m in diameter and contained six Piper individuals, limited by the number of cuttings available. Plants were cultivated without fertilizer or irrigation.

The interspecific richness treatment had three levels at all sites except Uaimii, where there were only two levels due to limited numbers of cuttings. At each site a subset of naturally occurring Piper species was selected to act as a species pool for each experimental plot (Table S3). These species were selected to constitute a breadth of genetic and functional diversity representative of the Piper community present at each site. Combinations of species in each plot for each level of the interspecific richness treatment were selected randomly from these species pools. The first level was a monoculture of Piper consisting of individuals of a single species. In the second level, two Piper species were randomly selected for planting (e.g. six individuals of each species). These species combinations were the same when crossed with the low and high levels of the intraspecific richness treatment (see below). In the high interspecific richness plots, the number of species varied between sites and included all Piper species available to produce cuttings (Table S1–Table S2). The minimum number of Piper species in a high richness plot was three in Uaimii, and the maximum was 12 in Costa Rica and Ecuador. Due to the low number of species available in Uaimii no two-species plots were planted. The interspecific richness treatment was replicated three times per site. In Mogi-Guaçu and Uaimii, each replicate contained a monoculture plot for each species.

An intraspecific richness treatment with two levels (clones taken from a single mother plant or individual cuttings all taken from unique mother plants) was crossed with the interspecific richness treatment. In Costa Rica and Ecuador, the high interspecific richness plots only included individuals from unique mother plants because 12 different species were planted. In Mogi-Guaçu and Uaimii, the experiments were fully crossed. In the high intraspecific richness treatment, cuttings were from different Piper individuals growing at least 10 m apart, so they were not from the same root system are were most likely genetically unique individuals.

A water addition treatment was crossed with the inter- and intraspecific richness treatments in Costa Rica, Ecuador, and Peru in order to examine how IPCC predicted regional mid-century increases in precipitation in these regions (38) may affect herbivory and trophic interactions. Plants were cultivated in five to six liter pots with drainage holes. Each potted plant under the water addition treatment was watered with an extra 2 L of water twice per month in Costa Rica and Ecuador. In Peru, 2 L of water were added to plants every three months. Water was rapidly applied as a flooding event to completely saturate the soil. All low richness plantings in Peru were watered, but in Costa Rica and Ecuador the low richness plantings were not watered, so only two levels of the species richness treatment were crossed with the water treatment (Figs. 1 & S1 A). Plant cultivation periods lasted in excess of one year at all sites, and as such water application treatments were performed in both the wet season and dry season.

Experimental plots were randomly located in the study sites; in Mogi-Guaçu and Uaimii they were organized in three replicate blocks. There was a minimum of 20 m between plots in blocks and 100 m between blocks. Randomly located plots were separated by a minimum of 50 m. The plots were spread across a maximum of 1,500 m in Peru and a minimum of 275 m in Ecuador. Experimental periods lasted between 1.4 and 2.8 years, depending on site (Table S2), and all measurements of herbivory were conducted on leaves that were initiated during these periods. Cuttings were replaced if they died during the first three months of the experiment. One of the authors was present at all of the sites for multiple visits to ensure as much standardization of treatments as possible.

To determine the effects of site variation in natural levels of precipitation on the outcomes of the water treatments, the absolute level of precipitation and precipitation anomaly relative to climate normals was collected for each month of the experimental periods in Costa Rica, Ecuador, and Peru (Fig. S2). Climate data for Costa Rica were obtained from La Selva Biological Station. Data for Ecuador and Peru were obtained using TerraClimate (39), and an interpolation error in the precipitation for Peru in February 2016, was corrected using data from the National Service of Meteorology and Hydrology of Peru (SENAMHI).

Measures of herbivory and insect richness

Herbivory was recorded by taking photographs of all the experimental leaves at the end of the experimental period. Additional photos were taken every three months in Uaimii and Mogi-Guaçu, and in the first five months of the experiment in Costa Rica. These data were used to measure herbivory and to determine the types of herbivores feeding on leaves based on patterns of damage. When herbivores were observed on plants, they were photographed but were left to continue feeding so as to not interfere with the experiment. Herbivory was quantified by eye for each type of herbivore on each leaf by a single parataxonomist with extensive experience measuring herbivory on Piper following established protocols (31). The amount of leaf area consumed was measured in function to the total leaf area by visually dividing the leaf into equal sized segments to determine the percent area missing. This was measured as a continuous value to the greatest possible accuracy. Insect herbivores were identified to the lowest taxonomic level possible based on their damage patterns (genus for specialist Lepidoptera, family for other Lepidoptera, family for Coleoptera, order for Orthoptera). Direct observations of herbivores were rare, so only damage patterns were used in analyses. The different types of damage recorded were used as proxy for insect insect richness on Piper. Hereafter, the term “insect richness” refers to the richness of herbivore damage patterns on plants.

Data analyses

The percentage of leaf tissue consumed by each insect taxon (based on damage patterns) on individual leaves was summed to determine the total percentage of herbivory on each leaf, and the variance in herbivory was calculated as the variance in leaf damage within each plant in a plot. Due to high mortality rates across sites, Piper interspecific richness in each plot was the final number of Piper species present in each plot at the end of the experiment (rather than the number of species planted) and was analyzed as a continuous covariate rather than a categorical treatment. Because most plant deaths occurred early in the experiments, final Piper interspecific richness more accurately reflects the local plant richness experienced by herbivores. In order to compare the magnitude of effect of intra and interspecific diversity on predictor variables, interspecific diversity for each plot was analyzed as the proportion of the interspecific diversity present in the species pools at each site. The effect of intra and interspecific diversity on each response variable is reported as the change in response as diversity changes from the lowest possible value at each site to the highest.

The effects of intraspecific richness, interspecific richness, and water addition on percent leaf area consumed, the percentage of damaged leaves, variance in herbivory, and insect insect richness were analyzed across and within sites using hierarchical Bayesian models (HBM). This framework acts as the Bayesian equivalent of a random-effects model where site is a random effect, allowing for generalized parameter estimates across sites. Analyses testing Piper intraspecific richness and interspecific richness were conducted using data from all sites; the effects of water addition and its interactions with richness were conducted using data from Ecuador, Costa Rica, and Peru. Due to high mortality in Peru, interactions between the water addition and intraspecific richness treatments could not be modeled for measures of herbivory or insect richness and were only tested for survival.

We acknowledge that treatment level combinations were not the same across the different sites, but that is the nature of random-effects models, where different levels of random factors, such as site or year, consist of treatment levels that are unique to that site or year. In fact, this type of design goes back to the origins of mixed models (40, 41), and are also a strength of the design, since lack of interactions between fixed and random effects increases generality (42). Even at a single site, manipulated variables in ecological experiments do not even consist of the same manipulations across all of the units of replication, as they suffer from problems such as multiple versions of treatments, interference, and noncompliance (43)

Bayesian structural equation models (BSEMs) were constructed for each site using all treatments as exogenous variables and insect richness and herbivory as endogenous variables. Three path models (Table S1) were developed under the assumptions that 1) intraspecific and interspecific richness may influence herbivory both directly and indirectly by modulating insect richness 2) water addition may influence both herbivory and insect richness and 3) insect richness may influence herbivory directly (Table S1). Models I, II, and III were tested in Ecuador, Costa Rica and Peru, while model III was used to analyze data from Mogi-Guaçu and Uaimii. Additional models incorporating interactions between intraspecific richness, interspecific richness, and water availability were tested in Costa Rica, Ecuador, and Peru.

Models were run at the leaf level for herbivory and insect richness, and at the plant level for the percentage of leaves with damage, variance of herbivory, and Piper survival. For all HBMs and BSEMs, model convergence was estimated visually using traceplots and an R^ discriminatory threshold of 1.1 (44). Model fit was determined via posterior predictive checks (PPC) using the χ2 discrepancy of the residuals as the discriminatory function (45). A PPC near 0.5 indicates a high model fit, while values near one or zero indicate poor fit. BSEM models were further compared using the Deviance Information Criterion (DIC).

All Bayesian models were written in JAGS via the jagsUI package in R (46) using Markov chain Monte Carlo (MCMC) samplers with weakly-informative priors. For the majority of models, MCMC runs were conducted for 10,000 iterations using the first 1,000 iterations as a burn-in phase to generate posterior distributions of parameter estimates for each response variable. HBMs modeling interactions required 20,000 iterations with the first 5,000 as burn-ins for all models to consistently converge. Mean parameter estimates and 95% credibility intervals (CIs) were calculated for all responses. 95% CIs which do not cross the y-axis are typically associated with less than a 2.5% type S error rate (47). A posterior probability of direction (PD) was calculated based on the percentage of the posterior distribution responding in the same direction as the median response. A PD of 95%, for example, indicates that the same direction of response (e.g. a positive or negative response) was observed in 95% of iterations, regardless of the magnitude of the response (48). PDs less than 95%, indicate lower confidence that a relationship exists, but can still be interpreted as the probability that an effect exists.

To analyze the effects of the experimental treatments on plant survival, survivorship curves were constructed in all sites. Analyses of Piper survival were based on the initially planted interspecific richness treatment of each plot because mortality occurred early in the experiment. The effect of water addition, intra- and interspecific richness and species identity on Piper survival were analyzed using Cox proportional hazard models. All data were analyzed using R software v4.1.0 (49).

Results

Considerable variation in herbivory and plant mortality was observed among study sites. Percent herbivory was lowest at Mogi; only 9% of leaf tissue was consumed by herbivores compared to 22% of leaf tissue consumed in Uaimii. Piper mortality was highest in Peru (89%), likely due to an El Niño event, and was lowest in Ecuador (27%; Fig. S3 A). Piper mortality resulted in the reduction of species richness in many plots and the loss of some combinations of treatments, notably the loss of all Piper in unwatered, low intraspecific richness plots in Peru (Fig. S1 B). Plots originally planted as high interspecific richness treatments had a final richness of between 1 and 12 species in Costa Rica, 9 and 11 species in Ecuador, 1 and 3 species in Peru, 2 and 4 species in Mogi, and only 2 species in Uaimii. We identified herbivore damage from 13 insect taxa at the five study sites, as well as damage by leaf miners of unknown orders. A total of 10 taxa were observed in Costa Rica, Ecuador, and Peru, 9 taxa were observed in Mogi, and 8 taxa were observed in Uaimii (Fig. S3 B). The majority of taxa in Costa Rica and Ecuador were Piper specialists, but the proportion of leaf tissue consumed by specialists was only greater than generalist damage in Costa Rica.

Our experiments revealed pronounced heterogeneity in ecosystem responses to changes in water and Piper diversity between sites (Figs. S4 & S5). PPCs for all HBMs were within 0.03 of 0.5, indicating models fit well. Model II (Table S1) was selected as the most parsimonious causal model for BSEMs in Costa Rica (PPC = 0.499), Ecuador (PPC = 0.499), and Peru (PPC = 0.498). Fit was high for model III in Mogi (PPC = 0.5) and Uaimii (PPC = 0.497). When all sites were analyzed together, percent herbivory was 4.2±3.6% (mean ± 95% CI) lower in plots that received additional water (PD = 98.7%), but we did not detect a cross site effect of water on the percentage of leaves with damage (Fig. 2 A-B). We did not detect any direct cross-site effects of water addition, insect richness, or plant diversity on the variance in herbivory (Fig. 2 C), or Piper survivorship (Fig. 2 D). Notably, greater Piper interspecific richness was associated with a 15 ± 18.6% increase in the richness of insect herbivores when comparing high diversity plots to low diversity plots across sites (PD = 95.0%; Fig. 2 E) and an indirect increase in herbivory was mediated by insect richness (Fig. 3). Insect richness was associated with an 8.8±2.8% increase in herbivory per insect taxon present (PD = 100%) and a 6.7±6.9% increase in the percentage of leaves with any damage (PP = 97%) (Fig. 2 A, B). Changes in both intraspecific richness and interspecific richness altered these ecosystem processes considerably, although effects varied in strength and direction across sites (Fig. 3, Table S4). Although intraspecific richness was not quantified, the genetic origin of Piper plants (clones vs. independent propagules) affected herbivory, plant mortality, and insect richness with effect sizes of similar or even greater magnitude than interspecific richness when standardized as the proportion of available species richenss at each site. However, we found that intra- and interspecific richness often had contrasting directions of effect on insect richness and multiple measures of herbivore pressure (Figs. 3, S4 B-C & S5 B-C). For example, in Costa Rica insect richness was 6.0 ± 5.9% (PD = 93.7%) lower in plots with high interspecific plant richness compared to low, while high intraspecific richness increased insect richness by 7.2 ± 6.9% (PD = 98.1%). In contrast, in Ecuador high interspecific richness was associated with a 43.6 ± 6.4% (PD = 100%) increase in insect richness, and high intraspecific richness decreased insect richness by 13.6 ± 5.2% (PD = 100%; Fig. S5 B-C).

Hierarchical Bayesian model parameter estimates for the effects of water availability intraspecific and interspecific Piper richness on herbivory, herbivorous insect richness, and Piper survivorship across sites Violins represent the cross-site posterior parameter distribution for each relationship in site level hierarchical Bayesian models.

Black lines represent the median posterior estimate and white bars represent 95% credibility intervals. Percentages above violins indicate the probability of an effect being positive or negative (as indicated by a negative probability) in response to an increase of the independent variable. Distributions for water addition compare watered and control plots; distributions for interspecific richness compare Piper species richness standardized as the proportion of the maximum richness used at a site; distributions for intraspecific richness compare low and high intraspecific richness treatments; distributions for insect richness compare responses per insect taxon present on an individual leaf

Direct and indirect effects of plant diversity and water availability on insect herbivores at five study sites.

Bayesian structural equation models comparing effects of different drivers of herbivorous insect richness and herbivory at five sites. Standardized path coefficients are means of the posterior distribution for the effects estimated at each causal path. Positive relationships are indicated in blue with triangular heads, and negative relationships are indicated in red with circular heads. Black arrows indicate path coefficients of zero magnitude. Dot plots summarize the standardized mean of the posterior distribution for each causal path with 95% and 80% credibility intervals. Asterisks indicate causal paths where the probability of an effect being positive or negative is >95%.

The effects of Piper richness were especially relevant when examining interactions between Piper richness and water addition (Fig. 4). The effects of water addition were altered by Piper intra- and interspecific richness at all sites. While a reduction in herbivory driven by water availability was observed in Costa Rica (4.7±2.5%, PD = 100%) and Peru (5.1±3.7%, PD = 100%), this effect was only present in Ecuador when intraspecific richness was high (Fig. S6). Across sites, water addition had negligible or negative effects on insect richness at low interspecific richness, but this effect was reduced or reversed when interspecific richness was high. Insect richness increased by 20.8±18.9% more in interspecifically diverse plots that received water compared to unwatered plots in Costa Rica (PD = 99%), 15.2±16.8% more in rich, watered plots in Ecuador (PD = 96%), and 29.3±31.5% more in Peru (PD = 98%) Additionally, Water addition had a negligible effect on insect richness in Ecuador when intraspecific richness was low, but increased insect richness by 12.1% when intraspecific richness was high (PD = 99%) (Fig. 4 B). Water addition had a negative effect on Piper survival in Costa Rica when intraspecific richness was low, but improved survival by 12.1 ± 8.5% in high intraspecific richness plots (PD = 99%; Fig. 4 C). Interactions with water in Peru may have been influenced by an El Niño event which resulted in abnormally low rainfall and high Piper mortality, while the typically wetter sites in Costa Rica and Ecuador experienced greater precipitation than average (Fig. S2). As there was 100% mortality in low intraspecific richness, unwatered plots in Peru, we could not explore interactions between intraspecific richness and water for any response except Piper mortality in this site.

Interactions between intraspecific or interspecific richness and water availability on herbivory, herbivorous insect richness, and Piper survival Bars indicate mean response and standard error of the mean.

Percentages indicate the probability that the two slopes are different as calculated using site level hierarchical Bayesian models. Due to high mortality in Peru, interactions between water and intraspecific richness could not be compared for any responses except mortality. Single and two species richness plots have been combined for visualization purposes only.

Cox analysis of plant survival

Water addition did not affect Piper survival in Costa Rica (z = -1.2, P > 0.2), or in Ecuador (z = 0.2, P > 0.8), but survival was reduced by 48% in plots without water in Peru (z = 3.2, P = 0.001). Intraspecific richness increased survival by 38% in Costa Rica (z = -3.3, P < 0.001) and by 32% in Ecuador (z = -4.9, P < 0.001), but had no effect on survival in Peru (z = - 1.55, P > 0.1), Mogi (z = -0.22, P > 0.8), or Uaimii (z = 0.18, P > 0.8). Interspecific richness had no effect on survival in Costa Rica (z = -1.5, P > 0.1), Ecuador (z = -0.56, P > 0.5), Peru (z = - 0.49, P > 0.6), or Uaimii (z = 0.21, P > 0.8). In Mogi, survival was reduced by 13% in plots with higher interspecific richness (z = -2.16, P = 0.031). There was an interaction between intraspecific richness and water addition in Costa Rica and Peru. Survival in Costa Rica increased in response to water in high intraspecific richness plots, and decreased in response to water in low intraspecific richness plots (z = 3.8, P < 0.001), while the opposite pattern was observed in Peru (z = -2.4, P = 0.02; Fig. S7). Although plant die-offs caused plot level species extirpation at several sites, plant species identity was not related to survival rates with the exception of P. peltatum, which had the lowest survival rate of any species planted in Costa Rica. Further statistical results are available in the supplement.

Discussion

Our results demonstrate two key patterns. First, the strength of effects of intraspecific richness on higher trophic levels is comparable to that of interspecific richness, supporting our predictions and corroborating recent studies demonstrating the importance of intraspecific richness (18, 19). However, the effects of intra- and interspecific richness on herbivory, insect richness, and plant survival can differ dramatically, in contrast to our predictions. Second, we found that perturbations in water availability can have complex effects on herbivores and plant survival, and that these effects can be modulated by plant diversity in a context sensitive manner. While our prediction that water availability would influence herbivory across sites was supported, our results suggest that biodiversity loss and climatic perturbations may have dramatically different effects on ecosystem function at local scales, which may diminish our ability to predict how local communities will change as anthropogenic stressors increase.

Although experimental methods varied somewhat between study sites, we do not believe this can fully explain the level of heterogeneity observed in ecosystem response. For example, the methods employed in Costa Rica and Ecuador were nearly identical, and yet the directions of effect of intra and interspecific diversity on insect richness were reversed in these sites (Fig. S5 B-C). The only meaningful differences between these two sites were the climate and the composition of the local Piper and insect communities. There was considerable variation in both biotic (Fig. S3 B) and abiotic factors (Table S2) across sites, which may have contributed to the heterogeneity observed in ecosystem response to Piper diversity.

As we did not directly measure plant stress or nutrition, it is difficult to determine the exact mechanism through which water addition reduced herbivory. The presence of an El Niño weather pattern during the course of the Peru experiment may have led the water addition treatment to relieve plants from drought stress, while water addition in Costa Rica and Ecuador may have added to water stress in treated plants as these sites received an above average level of rainfall during the course of the experiment. Despite this, water addition consistently suppressed herbivory in Costa Rica, Ecuador, and Peru under natural levels of Piper diversity, suggesting that predicted increases in precipitation in the next century (38) will dramatically alter the flow of resources from primary producers. Although we were only able to record the richness of insect herbivory patterns, this measure is indicative of the functional diversity of insect herbivores on Piper and any changes to this value represent changes in interactions between Piper and higher trophic levels (31). As such, the additional reduction in effects of water addition on insect richness when Piper richness was low suggest that biodiversity loss in tropical systems will alter the ability of higher trophic levels to respond to environmental perturbations.

As herbivore taxa can be differently affected by manipulations of diversity (50), variation in the direction of the effect of intra- and interspecific richness may be due in part to variation in the composition of insect communities and herbivore pressure measured across sites. Changes in neighborhood effects when small numbers of plant species dominate a community can lead either to the reduction or increase of herbivore pressure, dependent both on the nature of plant species lost and herbivore species present in a given community (51). As such, local variation in community composition has the potential to greatly alter the effects of both climate change and biodiversity loss on ecosystem function. Regardless of how biodiversity loss effects ecosystem function at large scales, variation in abiotic and biotic factors at locals scales can alter these effects, reducing our ability to predict how anthropogenic activity will alter ecosystem function. This is especially relevant in tropical systems, which have been the subject of far fewer studies of ecosystem function. As such, our knowledge of local scale effects on the relationship between biodiversity and ecosystem function remains limited in these systems.

A long-standing question in ecology has been the extent to which ecosystem function increases with biodiversity and if this relationship plateaus at a level past which ecological redundancy predominates. Recent results from less complex temperate grassland systems suggest that these ecosystems can be described by a mostly linear relationship between richness and function, where even rare species contribute uniquely to ecosystem function (52). In these systems, high contingency can be expected, where ecosystem-level effects will depend on most of the interacting species. In contrast, we might expect that diverse, tropical communities could be characterized by greater ecological redundancy and thus be subject to less contingency (53, 54). Despite these expectations, our results demonstrate heterogeneity in ecosystem response to changes in both intraspecific and interspecific richness in five tropical sites, suggesting that complexity in these systems may not reduce the contingency effects of biodiversity loss. Understanding the impact of biodiversity loss in tropical forests is fundamental to our ability to conserve those systems, and our findings highlight the importance of approaching the study of changes in ecosystem function as context sensitive responses in complex ecosystems.

eLife Assessment

This valuable study presents a large experimental study to examine the effects of plant species richness, plant genotypic richness, and soil water availability on herbivory patterns on Piper species in several tropical sites. The authors find that water availability and both intra- and interspecific plant diversity influence herbivory and herbivore diversity, but that the effects differ in significant ways across sites. The major weakness lies in findings that can be better presented in a less complicated fashion. The major claims are currently supported by a combination of solid and incomplete evidence, and the manuscript would benefit from better organization.

Reviewer #1 (Public Review):

This study reports a long-term, multisite study of tropical herbivory on Piper plants. The results are clear that lack of water leads to lower plant survival and altered herbivory. The results varied substantially among sites. The caveats are that ecosystem processes beyond water availability are not investigated although they are brought into play in the title and in the paper, that herbivory beyond leaf damage was not reported (there might be none, the reader needs to be shown the evidence for this), that herbivore diversity is defined by leaf damage (authors need to give evidence that this is a valid inference), that the plots were isolated from herbivores beyond their borders, that the effects of extreme climate events were isolated to Peru, that intraspecific variation in the host plants needs to be explained and interpreted in more detail, the results as reported are extremely complicated, the discussion is overly long and diffuse.

Reviewer #2 (Public Review):

This is an important and large experimental study examining the effects of plant species richness, plant genotypic richness, and soil water availability on herbivory patterns on Piper species in tropical forests.

A major strength is the size of the study and the fact that it tackled so many potentially important factors simultaneously. The authors examined both interspecific plant diversity and intraspecific plant diversity. They crossed that with a water availability treatment. And they repeated the experiment across five geographically separated sites.

The authors find that both water availability and plant diversity, intraspecific and interspecific, influence herbivore diversity and herbivory, but that the effects differ in important ways across sites. I found the study to be solid and the results to be very convincing. The results will help the field grapple with the importance of environmental change and biodiversity loss and how they structure communities and alter species interactions.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

We thank reviewer #1 for identifying the major caveats of the paper, and have split them out into separate comments below to address them.

Comment 1) The caveats are that ecosystem processes beyond water availability are not investigated although they are brought into play in the title and in the paper

Author response: We disagree that water availability is the only ecosystem process investigated in this study, as herbivory, plant mortality, and the maintenance of diversity in higher trophic levels are important processes within ecosystems. We have added text to the abstract and introduction clarifying that we consider these response measures to be ecosystem processes. Further language to this effect already exists in the abstract, methods, and discussion.

Comment 2) That herbivory beyond leaf damage was not reported (there might be none, the reader needs to be shown the evidence for this)

Author response: This is typically how herbivory is assessed in ecological studies, and our focus is on folivores. There may be additional herbivory in the form of fluid-sucking insects, shoot/root herbivory, etc., but these were not assessed. It would be interesting to assess these other forms of herbivory to see if they respond similarly with additional studies.

Comment 3) That herbivore diversity is defined by leaf damage (authors need to give evidence that this is a valid inference)

Author response: We thank reviewer #1 for pointing out the lack of written support for this claim. We have modified the methods (lines 138-139; 214-217) to clarify that this is a useful proxy for insect richness in the Piper system, and have added citations demonstrating it has been found to correlate well with insect richness in tropical forests.

Comment 4) That the plots were isolated from herbivores beyond their borders

Author response: This was not an assumption of the study. We have modified the methods (line 200) to make this clearer to the reader.

Comment 5) That the effects of extreme climate events were isolated to Peru

Author response: This was not an assumption of the study, rather it is an observation. While we consider it important to include observed climate differences between sites in the interpretation of our results, it was not necessary for there to be extreme climate events at other sites as we consider manipulated water availability to represent changes in precipitation that are expected to occur at these sites with climate change.

Comment 6) That intraspecific variation in the host plants needs to be explained and interpreted in more detail

Author response: We thank reviewer #1 for identifying that our current explanations needed development. We have modified the introduction to explore potential mechanisms relating intraspecific diversity to ecosystem function based on recent studies, and have modified the discussion to bring focus to why the effects of intraspecific differ from interspecific.

Reviewer #1 (Recommendations For The Authors):

Comment 1) Pare this material down to simpler results. The most significant to me is the intraspecific variation in damage. Were this broken out and reported in some detail it could be quite interesting. I find the results to be a confusing blizzard of multiple factors that differ among sites; after reading the paper twice I could not recall the takeaway lesson beyond that drought wrecks the diversity of herbivores and sometimes even kills the host plant.

Author response: We agree that the results are complicated given the variation in effects among sites, but this variation and complexity is important – and is in itself is one of the takeaway points. Unfortunately, nature is not simple. We have made several large edits to the results section, including the removal of methodological and otherwise redundant information, to hopefully bring the major takeaways into focus.

Reviewer #2 (Public Review):

Comment 1) This is an important and large experimental study examining the effects of plant species richness, plant genotypic richness, and soil water availability on herbivory patterns on Piper species in tropical forests.

A major strength is the size of the study and the fact that it tackled so many potentially important factors simultaneously. The authors examined both interspecific plant diversity and intraspecific plant diversity. They crossed that with a water availability treatment. And they repeated the experiment across five geographically separated sites.

The authors find that both water availability and plant diversity, intraspecific and interspecific, influence herbivore diversity and herbivory, but that the effects differ in important ways across sites. I found the study to be solid and the results to be very convincing. The results will help the field grapple with the importance of environmental change and biodiversity loss and how they structure communities and alter species interactions.

Author response: We thank reviewer #2 for their kind words.

Reviewer #2 (Recommendations For The Authors):

Comment 1) I was confused about why the authors measured species diversity/richness as a proportion of the species pool. This means that the metric of richness decreases if species are added to the species pool but not the plot/experiment. I think I understand it, but I suggest the authors explain this choice.

Author response: We thank reviewer #2 for pointing out that this was confusing. We have clarified the methods (lines 228-232) to explain that this choice was made to allow easier comparison between intra- and interspecific richness.

Comment 2) One of the stronger estimated relationships was a positive effect of plant species richness on insect richness. I found it a little hard to interpret this relationship. Is this just because there are host species specialists? So, with more host species there are more herbivore species? Or does insect richness increase multiplicatively with increasing plant species richness? One way to look for this would be for the authors to examine the relationship between plant species richness and the average number of herbivore damage types per plant species.

Author response: We agree that this is important for the reader to understand and have added text to the introduction and discussion sections explaining that this is the expectation based on theory and other empirical studies. We have additionally added text to the discussion (lines 386-388) pointing out that this pattern was not observed at all sites. While we agree that it would be interesting to explore if this effect was additive or multiplicative, we do not believe this is in the scope of the paper due to the methods used to measure insect richness.

Comment 3) Unless I missed it, some important information about the models was missing. E.g., what distributions were assumed for each of the variables? Any transformations?

Author response: We thank reviewer #2 for pointing this out, this information has been added to the methods (lines 272-274)

Comment 4) Why is there no model with water addition affecting insect richness directly but not percent herbivory directly?

Author response: While we originally decided to not include this model due to lack of theoretical support and low statistical performance, we have added references to this model (now model II) in the methods and results for consistency and to make model performance clearer to the reader. We have additionally moved supplemental table S1 to the main text to make the models and hypotheses tested by each model more accessible.

Comment 5) Fig. 2. What are the percentages above the figures? Maybe PD values?

Author response: These values are now clarified in the figure caption

Comment 6) L364 "can differ dramatically" This is vague and confusing. Differ in what way? From each other? Did the authors really expect plant richness to have the same effect on herbivory and plant survival? What would it mean anyway for plant richness to have the same effect on herbivory and plant survival?

Author response: We agree that the language here is confusing and thank reviewer #1 for drawing our attention to it. We have modified the discussion (lines 363-365) to clarify that the direction of effect of intraspecific richness can vary from the direction of effect of interspecific richness, rather than the effects on different response variables varying from each other.

Comment 7) L 375 "only meaningful differences" This statement feels a little overly strong. It seems like there is a good argument for this, but there could be other things going on.

Author response: We agree that the language here was unnecessarily strong, and have modified the discussion (lines 398-403) to focus on the lack of difference between methodologies at these two sites, and the observed differences in climate and community structure at each site.

Version 2

Abstract

Abstract

Declines in biodiversity generated by anthropogenic stressors at both species and population levels can alter emergent processes instrumental to ecosystem function and resilience. As such, understanding the role of biodiversity in ecosystem function and its response to climate perturbation is increasingly important, especially in tropical systems where responses to changes in biodiversity are less predictable and more challenging to assess experimentally. Using large scale transplant experiments conducted at five neotropical sites, we documented the impacts of changes in intraspecific and interspecific plant richness in the genus Piper on insect herbivory, insect richness, and ecosystem resilience to perturbations in water availability. We found that reductions of both intraspecific and interspecific Piper diversity had measurable and site specific effects on herbivory, herbivorous insect richness, and plant mortality. The responses of these ecosystem-relevant processes to reduced intraspecific Piper richness were often similar in magnitude to the effects of reduced interspecific richness. Increased water availability reduced herbivory by 4.2% overall, and the response of herbivorous insect richness and herbivory to water availability were altered by both intra and interspecific richness in a site dependent manner. Our results underscore the role of intraspecific and interspecific richness as foundations of ecosystem function and the importance of community and location specific contingencies in controlling function in complex tropical systems.

Introduction

As climate change and anthropogenic activity alter ecosystems at unprecedented rates, it has become critical to understand the consequences of biodiversity loss on ecosystem processes and the maintenance of ecosystem processes through species interactions. A complex mix of anthropogenic forces are eroding multiple dimensions of global biological diversity, including plant intraspecific, interspecific, and functional diversity (1,2). Plant diversity affects herbivore abundance and diversity, thereby influencing biomass allocation and energy fluxes between trophic levels (3). Because losses of plant diversity can destabilize the flow of resources to higher trophic levels (3,4), understanding the connection between biodiversity and trophic interactions is necessary to predict the consequences of the loss of primary producer biodiversity on ecosystem traits such as resilience to environmental perturbation (5,6). While climate change is associated with increased absolute precipitation in some regions and decreased precipitation in others, IPCC models predict mid-century increases in the frequency of extreme precipitation events in Central and South America (7–9), a phenomenon already observable in many ecosystems (10). Measuring how precipitation change will affect the relationship between biodiversity and ecosystem function is therefore increasingly important, particularly in diverse tropical systems.

Very few multi-site, manipulative diversity experiments have been reported from tropical areas compared to temperate environments (11), limiting our knowledge of the role of biodiversity in ecosystem function in the most species-rich regions of the planet (12). In the context of several established hypotheses (Table 1) we investigate how multiple dimensions of plant diversity affect ecosystem processes at five neotropical sites and explore how diversity modulates how ecosystems respond to changes in water availability at three of those sites. We focus on ecosystem responses that represent changes in energy fluxes between trophic levels as measured by herbivory, herbivore diversity and plant mortality. By altering plant uptake of nutrients and plant defense production, abnormal levels of precipitation can alter herbivore pressure, affecting the movement of resources into higher trophic levels (13). As such, extreme dry and wet periods of climate are expected to strongly perturb plant-insect interactions and thereby alter ecosystem function (13–15). The insurance hypothesis suggests that greater biodiversity can act to stabilize ecosystems and improve their resilience to environmental change (4,16). While greater interspecific plant richness is expected to lead to increased diversity in higher trophic levels due to the accumulation of specialist herbivores, field studies have demonstrated both positive and neutral effects of interspecific plant richness on ecosystem resilience (17,18). Despite traditional views that interspecific richness has a greater impact on ecosystem processes than intraspecific diversity (19), recent research suggests that plant intra- and interspecific richness can have similar effects on ecosystem productivity and consumer abundance (20,21). As changes to intraspecific richness can alter the diversity of resources available to herbivores (22), it is necessary to consider both inter and intraspecific diversity when investigating the effects of biodiversity loss on ecosystem processes.

We conducted common garden experiments at five sites across Central and South America to test the insurance hypothesis by quantifying 1) the relative strength of intra- and interspecific plant richness in driving ecosystem function and 2), the effects of increased water availability on ecosystem function. Using 33 species in the genus Piper (Piperaceae) as a model system (23), we manipulated intra- and interspecific plant richness in Costa Rica, Ecuador, Peru, and two sites in Brazil. We additionally manipulated water availability in Costa Rica, Ecuador, and Peru (Figs. 1 & S1 A). We predicted that reduced Piper diversity would lead to reduced diversity of higher trophic levels, that water addition would lead to altered herbivore pressure, and that lower Piper diversity would be associated with more extreme changes in herbivory and plant mortality in response to water addition. Finally, we predicted that changes in intraspecific and interspecific plant richness would affect ecosystem processes, including herbivory, herbivore diversity, and plant mortality, with similar magnitudes.

Treatments of intraspecific richness, interspecific richness, and water addition used in each of the five study sites.

White tiles represent treatment combinations which were not tested in a given site

Methods

Study sites and focal plant genus

We conducted a large-scale transplant experiment replicated across five sites spanning 42° latitude in the Neotropics (Fig. 1 & Table S1) encompassing a range in annual precipitation from 1271 mm to 4495 mm (Table S1). At each site, we studied herbivory on planted individuals in the genus Piper (Piperaceae) in response to experimental treatments. Study sites included lowland equatorial humid forest at La Selva Biological Station, Costa Rica; high elevation equatorial humid forest in Yanayacu Biological Station, Ecuador; high elevation equatorial humid forest at El Fundo Génova, Peru; lowland seasonally dry gallery forest in the cerrado within the phytogeographic domain of the Atlantic Forest in Mogi-Guaçu Biological Reserve, Brazil; and lowland seasonally semideciduous forest in the transition between the Atlantic Forest and the cerrado phytogeographic domains in Uaimii State Forest, Brazil. Climate classifications follow the Köppen-Geiger climate model (Table S1; 22).

Multiple species of Piper are found at all sites, ranging from 11 species in Mogi-Guaçu to 50 species in Costa Rica (25). Piper is an ideal genus for large-scale comparative studies, as it is found across the Neotropics and subtropics, from ∼10° N to about ∼32° S. In addition to being widespread, Piper is abundant and diverse across its range, encompassing ∼1000 species in the Neotropics (26). Piper has been the subject of detailed studies of herbivory, and its herbivore fauna has been surveyed across its range (27–38). Members of the genus host both specialist caterpillars and beetles as well as generalist caterpillars and Orthopterans (23,27–30,32). Leaf damage patterns produced by the different classes of herbivores are well documented and allow for the determination of unique taxa of herbivores (32).

Experimental design

Executing experiments across the Americas presents challenges, including nuanced variations in methodologies at each site. Nonetheless, the advantages of this expansive and consistent approach provides a greater understanding in the role of biodiversity in ecosystem function than examining isolated single-site studies. Here we describe the experimental design applied across sites, see Figures 1 & S1, and Tables S1 & S2 for site specific details.

At each study site, a factorial experiment was implemented to test the effects of plant interspecific and intraspecific richness on herbivory, variation in herbivory, and insect richness. Experimental plants were propagated from cuttings of naturally occurring Piper plants (typically with three nodes and zero to one leaves). At each site a subset of naturally occurring Piper species was selected to act as a species pool for each experimental plot (Table S2). These species were selected to constitute a breadth of genetic and functional diversity representative of the Piper community present at each site. In all sites except Uaimii, experimental plots measured 4 m in diameter and contained 12 Piper individuals, planted either in pots with locally derived soil (for experiments manipulating water availability) or directly into the ground (for experiments without water additions in Mogi-Guaçu and Uaimii; Table S1) and cultivated without fertilizer or irrigation. Interspecific richness treatment levels consisted of single species monoculture plots, two species plots, and high richness plots with the maximum number of species available to produce cuttings (Table S1). The minimum number of Piper species in a high richness plot was three in Uaimii, and the maximum was 12 in Costa Rica and Ecuador. In plots with more than one species, species were randomly sampled from the species pool for that site.

An intraspecific richness treatment was crossed with the interspecific richness treatment. Intraspecific richness was manipulated by clones taken from a single mother plant (low richness) or individual cuttings all taken from unique mother plants (high richness). In the high intraspecific richness treatment, cuttings were from different Piper individuals growing at least 10 m apart, to eliminate the likelihood of a shared root system thus representing genetically unique individuals. In Costa Rica and Ecuador, the high interspecific richness plots only included individuals from unique mother plants because all cuttings were taken from different species.

A water addition treatment was crossed with the inter- and intraspecific richness treatments in Costa Rica, Ecuador, and Peru in order to examine how IPCC predicted regional mid-century increases in precipitation in these regions (9) may affect herbivory and trophic interactions. Plants were cultivated in five to six liter pots with drainage holes. Each potted plant under the water addition treatment was watered with 2 L of water twice per month in Costa Rica and Ecuador. In Peru, 2 L of water were added to plants every three months. Water was rapidly applied as a flooding event to completely saturate the soil. Plant cultivation periods lasted in excess of one year at all sites, and as such water was added in both the wet season and dry season. Experimental plots were randomly located in the study sites; in Mogi-Guaçu and Uaimii they were organized in three replicate blocks. There was a minimum of 20 m between plots in blocks and 100 m between blocks. Randomly located plots were separated by a minimum of 50 m. Experimental periods lasted between 1.4 and 2.8 years, depending on the site (Table S1), and all measurements of herbivory were conducted on leaves that were initiated during these periods. Cuttings were replaced if they died during the first three months of the experiment. One of the authors was present at all of the sites for multiple visits to ensure as much standardization of treatment applications as possible.

Piper mortality resulted in a reduction in species richness in many plots and the loss of some treatment combinations, notably the loss of all Piper in unwatered, low intraspecific richness plots in Peru (Fig. S1B). Plots originally planted as high interspecific richness treatments had a final richness of between 1 and 12 species in Costa Rica, 9 and 11 species in Ecuador, 1 and 3 species in Peru, 2 and 4 species in Mogi, and only 2 species in Uaimii. To determine the effects of site variation in natural levels of precipitation on the outcomes of the water addition treatment, the absolute level of precipitation and precipitation anomalies relative to climate normals were collected for each month of the experimental periods in Costa Rica, Ecuador, and Peru (Fig. S2). Climate data for Costa Rica were obtained from La Selva Biological Station. Data for Ecuador and Peru were obtained using TerraClimate (39), and an interpolation error in the precipitation for Peru in February 2016, was corrected using data from the National Service of Meteorology and Hydrology of Peru (SENAMHI).

Measures of herbivory and insect richness

Plots were open to naturally occurring herbivores, and herbivory was recorded by taking photographs of all the experimental leaves at the end of the experimental period. Additional photos were taken every three months in Uaimii and Mogi-Guaçu, and in the first five months of the experiment in Costa Rica. These data were used to measure herbivory and to determine the types of herbivores feeding on leaves based on patterns of damage. When herbivores were observed on plants, they were photographed but were left to continue feeding so as to not interfere with the experiment. Herbivory was quantified by eye for each type of herbivore on each leaf by a single parataxonomist with extensive experience measuring herbivory on Piper following established protocols (32). The amount of leaf area consumed was measured in relation to the total leaf area by visually dividing the leaf into equal sized segments to determine the percent area missing. This was measured as a continuous value to the greatest possible accuracy, typically 1-5% of the total leaf area. Insect herbivores were identified to the lowest taxonomic level possible based on their damage patterns (genus for specialist Lepidoptera, family for generalist Lepidoptera, family for Coleoptera, order for Orthoptera). Direct observations of herbivores were rare, so only damage patterns were used in analyses. As insect damage patterns on leaves are tightly correlated with insect richness in tropical forests (40), the different types of damage recorded were used as proxy for the richness of above ground insect herbivores on Piper. Hereafter, the term “insect richness” refers to the richness of insect herbivore damage patterns on plants.

Data analyses

The percentage of leaf tissue consumed by each insect taxon (based on damage patterns) on individual leaves was summed to determine the total percentage of herbivory on each leaf, and the variance in herbivory was calculated as the variance in leaf damage within each plant in a plot. Due to high mortality rates across sites, Piper interspecific richness in each plot was the final number of Piper species present in each plot at the end of the experiment (rather than the number of species planted) and was analyzed as a continuous covariate rather than a categorical treatment. Because most plant deaths occurred early in the experiments, final Piper interspecific richness more accurately reflects the local plant richness experienced by herbivores. Interspecific diversity for each plot was analyzed as the proportion of the interspecific diversity present in the species pools at each site. This enabled easier comparisons between intra and interspecific richness, as intraspecific richness was only quantified as low and high, where high treatments represent the maximum intraspecific richness available at each site. The effect of intra and interspecific diversity on each response variable is reported as the change in response as diversity changes from the lowest possible value at each site to the highest.

The effects of intraspecific richness, interspecific richness, and water addition on percent leaf area consumed, the percentage of damaged leaves, variance in herbivory, and insect richness were analyzed across and within sites using hierarchical Bayesian models (HBM). This framework acts as the Bayesian equivalent of a random-effects model where site is a random effect, allowing for generalized parameter estimates across sites. Analyses testing Piper intraspecific richness and interspecific richness were conducted using data from all sites; the effects of water addition and its interactions with richness were conducted using data from Ecuador, Costa Rica, and Peru. Due to high mortality in Peru, interactions between the water addition and intraspecific richness treatments could not be modeled for measures of herbivory or insect richness, and were only tested for survival.

We acknowledge that treatment level combinations were not the same across the different sites, but experimental designs of this nature are encompassed within the framework of random-effects models, where different levels of random factors, such as site or year, consist of treatment levels that are unique to that site or year. This type of experimental design goes back to the origins of mixed models (41,42), and the lack of interactions between fixed and random effects increases generality in these models (43). Even at a single site, manipulated variables in ecological experiments do not even consist of the same manipulations across all of the units of replication, as they suffer from problems such as multiple versions of treatments, interference, and noncompliance (44)

Bayesian structural equation models (BSEMs) were constructed for each site using all treatments as exogenous variables and insect richness and herbivory as endogenous variables. Three path models were developed under the assumptions that 1) intraspecific and interspecific richness may influence herbivory both directly and indirectly by modulating insect richness 2) water addition may influence both herbivory and insect richness and 3) insect richness may influence herbivory directly (Table 1). Models I-III were tested in Ecuador, Costa Rica and Peru. while a model without the water addition variable was used to analyze data from Mogi-Guaçu and Uaimii. Additional models incorporating interactions between intraspecific richness, interspecific richness, and water availability were tested in Costa Rica, Ecuador, and Peru.

Models were run at the leaf level for herbivory and insect richness, and at the plant level for the percentage of leaves with damage, variance of herbivory, and Piper survival. For all HBMs and BSEMs, model convergence was estimated visually using traceplots and an RLJ discriminatory threshold of 1.1 (45). Model fit was determined via posterior predictive checks (PPC) using the sum of squares of the residuals as the discriminatory function (46). A PPC near 0.5 indicates a high model fit, while values near one or zero indicate poor fit. BSEM models were further compared using the Deviance Information Criterion (DIC).

All Bayesian models were written in JAGS via the jagsUI package in R (47) using Markov chain Monte Carlo (MCMC) sampling with weakly-informative priors. Residuals were modeled as normally distributed based on PPC comparisons between models. Models using gamma distributions for herbivory and binomial distributions for damage presence and mortality were found to consistently underestimate the magnitude of variance in the data based on PPC. For the majority of models, MCMC runs were conducted for 10,000 iterations using the first 1,000 iterations as a burn-in phase to generate posterior distributions of parameter estimates for each response variable. HBMs modeling interactions required 20,000 iterations with the first 5,000 as burn-ins for all models to consistently converge. Mean parameter estimates and 95% credible intervals (CIs) were calculated for all responses. 95% CIs which do not cross the y-axis are typically associated with less than a 2.5% type S error rate (48). A posterior probability of direction (PD) was calculated based on the percentage of the posterior distribution responding in the same direction as the median response. A PD of 95%, for example, indicates that the same direction of response (e.g. a positive or negative response) was observed in 95% of iterations, regardless of the magnitude of the response (49). PDs less than 95%, indicate lower confidence that a relationship exists, but can still be interpreted as the probability that an effect exists.

To analyze the effects of the experimental treatments on plant survival, survivorship curves were constructed for all sites. Analyses of Piper survival were based on the initially planted interspecific richness treatment of each plot because mortality occurred early in the experiment. The effects of water addition, intra- and interspecific richness and species identity on Piper survival were analyzed using Cox proportional hazard models. All data were analyzed using R software v4.1.0 (50).

Results

Considerable variation in herbivory and plant mortality was observed among study sites. Percent herbivory was lowest at Mogi where only 9% of leaf tissue was consumed by herbivores compared to 22% of leaf tissue consumed in Uaimii. Piper mortality was highest in Peru (89%), likely due to El Niño related drought, and was lowest in Ecuador (27%; Fig. S3A). We identified herbivore damage from 13 insect taxa at the five study sites, as well as damage by leaf miners of unknown orders. Damage from a total of 10 taxa were observed in Costa Rica, Ecuador, and Peru, 9 taxa were observed in Mogi, and 8 taxa were observed in Uaimii (Fig. S3 B). The majority of taxa in Peru, Mogi, and Uaimii were generalist herbivores, while 60% of taxa in Costa Rica and 58% of taxa in Ecuador were Piper specialists. The proportion of leaf tissue consumed by specialists was only greater than generalist damage in Costa Rica.

Our experiments revealed pronounced heterogeneity in ecosystem responses to water availability and Piper diversity between sites (Figs. S4 & S5). Posterior predictive checks (PPCs) for all hierarchical Bayesian models (HBMs), and for models I, and III were within 0.03 of 0.5, indicating models fit well. Model III (Table 1) was selected as the most parsimonious causal model for BSEMs in Costa Rica (PPC = 0.499), Ecuador (PPC = 0.499), and Peru (PPC = 0.498). Fit was high for models in Mogi (PPC = 0.5) and Uaimii (PPC = 0.497). Across all sites where the water addition treatment was applied, percent herbivory was 4.2±3.6% (mean ± 95% CI) lower in plots that received additional water (PD = 98.7%; Fig. 2 A-B). Greater Piper interspecific richness was associated with a 15±18.6% increase in the richness of insect herbivores (PD = 95.0%; Fig. 2E) and an indirect increase in herbivory was mediated by insect richness (Fig. 3). Insect richness was associated with an 8.8±2.8% increase in herbivory per insect taxon present (PD = 100%) and a 6.7±6.9% increase in the percentage of leaves with any damage (PP = 97%) (Figs. 2A, B). Intra- and interspecific richness affected herbivory, although effects varied in strength and direction across sites (Fig. 3, Table S3). Intraspecific richness had similar or greater effects on plant mortality and insect richness than interspecific richness. However, intra- and interspecific richness often had contrasting directions of effect on insect richness and measures of herbivore pressure (Figs. 3, S4B-C & S5B-C). For example, in Costa Rica insect richness was 6.0 ± 5.9% (PD = 93.7%) lower in plots with high interspecific plant richness, while high intraspecific richness increased insect richness by 7.2 ± 6.9% (PD = 98.1%). In contrast, in Ecuador high interspecific richness was associated with a 43.6 ±6.4% (PD = 100%) increase in insect richness, and high intraspecific richness decreased insect richness by 13.6 ±5.2% (PD = 100%; Fig. S5B-C).

Hierarchical Bayesian model parameter estimates for the effects of water availability, as well as intraspecific and interspecific Piper richness on herbivory, herbivorous insect richness, and Piper survivorship across sites

Violins represent the cross-site posterior parameter distribution for each relationship in site level hierarchical Bayesian models. Black lines represent the median posterior estimate and white bars represent 95% credible intervals. Percentages above violins indicate the probability of an effect being positive or negative (as indicated by a negative probability) in response to an increase of the independent variable. Distributions for water addition compare watered and control plots; distributions for interspecific richness compare Piper species richness standardized as the proportion of the maximum richness used at a site; distributions for intraspecific richness compare low and high intraspecific richness treatments; distributions for insect richness compare responses per insect taxon present on an individual leaf

Direct and indirect effects of plant diversity and water availability on insect herbivores at five study sites.

Bayesian structural equation models comparing effects of different drivers of herbivorous insect richness and herbivory at five sites. Standardized path coefficients are means of the posterior distribution for the effects estimated at each causal path. Positive relationships are indicated in blue with triangular heads, and negative relationships are indicated in red with circular heads. Black arrows indicate path coefficients of zero magnitude. Dot plots summarize the standardized mean of the posterior distribution for each causal path with 95% and 80% credible intervals. Asterisks indicate causal paths where the probability of an effect being positive or negative is >95%

The effects of water addition were altered by Piper intra- and interspecific richness at all sites (Fig. 4). Water availability reduced herbivory in Costa Rica (4.7 ±2.5%, PD = 100%) and Peru (5.1 ±3.7%, PD = 100%), but this effect was only present in Ecuador when intraspecific richness was high (Fig. S6). Across sites, water addition had negligible or negative effects on insect richness at low interspecific richness, but this pattern was reduced or reversed when interspecific richness was high. Insect richness increased by 20.8 ±18.9% more in interspecifically diverse plots water compared to unwatered plots in Costa Rica (PD = 99%). Insect richness increased by 15.2 ±16.8% more in rich, watered plots in Ecuador (PD = 96%), and 29.3 ±31.5% more in Peru (PD = 98%). Additionally, water addition had a negligible effect on insect richness in Ecuador when intraspecific richness was low, but increased insect richness by 12.1% when intraspecific richness was high (PD = 99%; Fig. 4B).

Interactions between intraspecific or interspecific richness and water availability on herbivory, herbivorous insect richness, and Piper survival

Bars indicate mean response and standard error of the mean. Percentages above each sub-figure indicate the probability that the two slopes are different as calculated using site level hierarchical Bayesian models. Due to high mortality in Peru, interactions between water and intraspecific richness could not be compared for any responses except mortality. Single and two species richness plots have been combined for visualization purposes only

Water addition had a negative effect on Piper survival in Costa Rica when intraspecific richness was low, but improved survival by 12.1 ±8.5% in high intraspecific richness plots (PD = 99%; Fig. 4C). Interactions with water in Peru may have been influenced by an El Niño related drought which resulted in high Piper mortality, while the typically wetter sites in Costa Rica and Ecuador experienced greater precipitation than average (Fig. S2).

Analysis of plant survival

Water addition did not affect Piper survival in Costa Rica (z = -1.2, P > 0.2), or in Ecuador (z = 0.2, P > 0.8), but survival was reduced by 48% in plots without water in Peru (z = 3.2, P = 0.001). Intraspecific richness increased survival by 38% in Costa Rica (z = -3.3, P < 0.001) and by 32% in Ecuador (z = -4.9, P < 0.001), but had no effect on survival in Peru (z = - 1.55, P > 0.1), Mogi (z = -0.22, P > 0.8), or Uaimii (z = 0.18, P > 0.8). Interspecific richness had no effect on survival in Costa Rica (z = -1.5, P > 0.1), Ecuador (z = -0.56, P > 0.5), Peru (z = - 0.49, P > 0.6), or Uaimii (z = 0.21, P > 0.8). In Mogi, survival was reduced by 13% in plots with higher interspecific richness (z = -2.16, P = 0.031). There was an interaction between intraspecific richness and water addition in Costa Rica and Peru. Survival in Costa Rica increased in response to water in high intraspecific richness plots, and decreased in response to water in low intraspecific richness plots (z = 3.8, P < 0.001), while the opposite pattern was observed in Peru (z = -2.4, P = 0.02; Fig. S7). Although plant die-offs cause a loss in richness in some plots, plant species identity was not related to survival with the exception of P. peltatum, which had the lowest survival rate of any species planted in Costa Rica. Further statistical results are available in the supplement.

Discussion

Our results demonstrate two key patterns. First, the strength of effects of intraspecific richness on higher trophic levels is comparable to that of interspecific richness, supporting our predictions and corroborating recent studies demonstrating the importance of intraspecific richness (20,21,51). However, the direction of the effect of intraspecific richness on herbivory, insect richness, and plant survival can vary dramatically from the direction of effect of interspecific richness, in contrast to our predictions. Second, we found that perturbations in water availability can have complex effects on herbivores and plant survival, and that these effects can be modulated by plant diversity in a context sensitive manner. While our prediction that water availability would influence herbivory across sites was supported, our results suggest that biodiversity loss and climatic perturbations may have dramatically different effects on ecosystem function at local scales, which may diminish our ability to predict how local communities will change as anthropogenic stressors increase.

As we did not directly measure plant stress or nutrition, it is difficult to determine the exact mechanism through which water addition reduced herbivory. The presence of an El Niño weather pattern during the course of the Peru experiment may have led the water addition treatment to relieve plants from drought stress, while water addition in Costa Rica and Ecuador may have added to water stress in treated plants as these sites received an above average level of rainfall during the course of the experiment. Despite this, water addition consistently suppressed herbivory in Costa Rica, Ecuador, and Peru under natural levels of Piper diversity, suggesting that predicted increases in precipitation in the next century (52) will dramatically alter the flow of resources from primary producers. Although we were only able to record the richness of insect herbivory patterns, this measure is indicative of the functional diversity of insect herbivores on Piper and changes to this value represent changes in interactions between Piper and higher trophic levels (32,40). As such, the additional reduction in effects of water addition on insect richness when Piper richness was low suggest that biodiversity loss in tropical systems will alter the ability of higher trophic levels to respond to environmental perturbations.

While our prediction that increased Piper interspecific richness would lead to increased insect diversity was met in the majority of sites, interspecific richness was associated with decreased insect diversity in Costa Rica and Peru. As herbivore taxa can be differently affected by manipulations of diversity (53), variation in the direction of the effect of intra- and interspecific richness may be due in part to variation in the composition of insect communities and herbivore pressure measured across sites. Changes in neighborhood effects when small numbers of plant species dominate a community can lead either to the reduction or increase of herbivore pressure, dependent both on the nature of plant species lost and herbivore species present in a given community (54). As such, local variation in community composition has the potential to greatly alter the effects of both climate change and biodiversity loss on ecosystem function.

Although experimental methods varied somewhat between study sites, this cannot fully explain the level of heterogeneity observed in ecosystem response. For example, the methods employed in Costa Rica and Ecuador were nearly identical, and yet the directions of effect of intra- and interspecific diversity on insect richness were reversed in these sites (Fig. S5 B-C). There was considerable variation in both biotic (Fig. S3 B) and abiotic factors (Table S1) across sites, which may have contributed to the heterogeneity observed in ecosystem response to Piper diversity. Regardless of how biodiversity loss effects ecosystem function at large scales, variation in abiotic and biotic factors at locals scales can alter these effects, reducing our ability to predict how anthropogenic activity will alter ecosystem function. This is especially relevant in tropical systems, which have been the subject of far fewer studies of ecosystem function than temperate ecosystems. As such, our knowledge of local effects on the relationship between biodiversity and ecosystem function remains limited in these systems.

A long-standing question in ecology has been the extent to which ecosystem function increases with biodiversity and if this relationship plateaus at a level past which ecological redundancy predominates. Recent results from less complex temperate grassland systems suggest that these ecosystems can be described by a mostly linear relationship between richness and function, where even rare species make unique contributions to ecosystem function (55). In these systems, high contingency can be expected, where ecosystem-level effects will depend on most of the interacting species. In contrast, we might expect that diverse, tropical communities could be characterized by greater ecological redundancy and thus be subject to less contingency (56,57). Despite these expectations, our results demonstrate heterogeneity in ecosystem response to changes in both intraspecific and interspecific richness in five tropical sites, suggesting that complexity in these systems may not reduce the contingency effects of biodiversity loss. Understanding the impact of biodiversity loss in tropical forests is fundamental to our ability to conserve those systems, and our findings highlight the importance of approaching the study of changes in ecosystem function as context sensitive responses in complex ecosystems.

Supporting information